---
title: "whale_argos_data_pipeline"
output: html_document
---





```{r warning=FALSE}

library(ggplot2)
library(lubridate)
library(argosfilter)
library(tidyr)
library(plyr)
library(dplyr)
library(tidyverse)
library(rnaturalearth)
library(plotly)
library(aniMotum)
library(trip)
library(viridis)


world_map <- map_data("world")%>%fortify()

sc <- scale_colour_gradientn(colours = viridis(100), limits=c(0,1))

```



```{r}
# Load new raw data including 2020, 2021, 2022 until 2023/6/6
# total of 57902 locs 

raw_argos_df <- list.files(path="D:/auckland/nzsrw/maps/maps/raw_argos_07102023",pattern=".csv",full.names = TRUE)
raw_argos_df<-lapply(raw_argos_df,read.csv)
raw_argos_df<-do.call("rbind",raw_argos_df)

#rename and remove columns we dont need 
raw_argos_df <- raw_argos_df[c("Ptt", "Date", "Quality", "Longitude", "Latitude")]
names(raw_argos_df) <- c("id", "date", "lc", "lon", "lat")

# change date format from hh:mm:ss dd-mm-yyyy to yyyy-mm-dd hh:mm:ss

dates <- strptime(raw_argos_df$date,format = "%H:%M:%S %d-%b-%Y")
dates_changed <- format(dates,"%Y-%m-%d %H:%M:%S")
raw_argos_df$date<- dates_changed

raw_argos_df$date <-ymd_hms(raw_argos_df$date, tz = "GMT")

remove(dates_changed)
remove(dates)

# Order the data by id and date
raw_argos_df <- raw_argos_df[order(raw_argos_df$id, raw_argos_df$date),]

# Take a quick peek 
head(raw_argos_df)
str(raw_argos_df)

# number of whales in the dataset
length(unique(raw_argos_df$id))
# how many loc fixes we have for each individual 
table(raw_argos_df$id)

```

```{r sda filter}

# apply sda filter with vmax = 25 m/s 
raw_argos_df <- ddply(raw_argos_df, ~id, function(d){
      d$argosfilter <- sdafilter(lat = d$lat, 
                               lon = d$lon, 
                               lc = d$lc, 
                               dtime = d$date, vmax = 25)
        return(d)})

#visualize 
                
ggplot()+
  geom_point(data=data.frame(raw_argos_df[raw_argos_df$argosfilter=="not",]),aes(lon,lat),col ="blue")+
  geom_point(data=data.frame(raw_argos_df[raw_argos_df$argosfilter=="removed",]),aes(lon,lat),col="red")+
  geom_polygon(data=world_map,aes(x=long,y=lat,group=group))+
  coord_fixed(xlim=c(160,170), ylim=c(-56,-45))+
  theme_bw()+
  theme(panel.grid=element_blank())

#ggplotly()


table(raw_argos_df$argosfilter)# removed n=5772 vmax=25

#Subset to exclude the erroneous locations

filtered_argos_df <- raw_argos_df %>% 
  filter(argosfilter != "removed") %>% 
  dplyr::select(-argosfilter)

#Who is left? How many positions?
filtered_argos_df %>% 
  group_by(id) %>% 
  dplyr::summarize(nb_locations = n())

#How does that compare to the raw dataset?
tab_1 <- raw_argos_df %>% 
  group_by(id) %>% 
  dplyr::summarize(nb_locations = n())

tab_2 <- filtered_argos_df %>% 
  group_by(id) %>% 
  dplyr::summarize(nb_locations = n())

tab  <- plyr::join(data.frame(tab_1), data.frame(tab_2), by="id")
colnames(tab) <- c("id", "raw_locs", "filt_locs")
tab

#52130 locs left 

```


```{r remove duplicate}

pre_dup <- nrow(filtered_argos_df) # to get the current number of data points

# create dummy variable
filtered_argos_df$index <- c(1:nrow(filtered_argos_df))

# run each tag in a loop check to check for duplicates
# if there is a time duplicate, select the best quality position or simply the first position
filtered_argos_df <- ddply(filtered_argos_df, ~id, function(d){
  toremove <- c()
  for (i in c(2:nrow(d))) {
    if (d$date[i] == d$date[i-1]) {
      dd <- d[(i-1):i,]
      r <- dd[dd$lc == ave(dd$lc, FUN = min), "index"] # find the lowest quality
      toremove <- c(toremove, r[1]) #select first element of r in case both locations have the same lq
    }
  }
  if (length(toremove) > 0){d <- d[!(d$index %in% toremove), ]}
  return(d)
})
# remove dummy variable
filtered_argos_df$index <- NULL
pre_dup - nrow(filtered_argos_df) # to get an understanding of how many duplicates were removed

# 1517 locations duplicate 
# 50613 locations left 

```

```{r time difference}

# Look at the time difference between locations

time_diff_hours_df <- ddply(filtered_argos_df, ~id, function(d){
  d$time_diff_hours <- NA
  for (i in 2:nrow(d)){
    d$time_diff_hours[i] = as.numeric(difftime(d$date[i], d$date[i-1], units = "hours"))}
  return(d)
})

# mean time difference between locations (in hours)
mts <- aggregate(time_diff_hours~id, time_diff_hours_df, mean)
mts #this is the mean time step

mxts <- aggregate(time_diff_hours~id, time_diff_hours_df, max)
mxts# this is the max time step

mnts <- aggregate(time_diff_hours~id, time_diff_hours_df, min)
mnts # this is the minimum time step

mets <- aggregate(time_diff_hours~id, time_diff_hours_df, median)
mets # this is the median time step

# track segments
trackseg_argos_df <- ddply(time_diff_hours_df, ~id, function(d){
ind <- which(d$time_diff_hours > 24)
d$mark <- 0
d$mark[ind] <- 1
d$track_seg <- cumsum(d$mark)
  return(d)
})

# Now create a new id based on track segment
trackseg_argos_df$track_id <- paste(trackseg_argos_df$id, "-", trackseg_argos_df$track_seg, sep="")

```

```{r}
# check overall
ggplot(trackseg_argos_df, aes(lon, lat)) +
  geom_point(size=0.5, aes(col = as.factor(track_id ))) +
  geom_polygon(data=world_map,aes(x=long,y=lat,group=group))+
  coord_fixed(xlim=c(160,180), ylim=c(-56,-45))+
  theme_bw()+
  theme(panel.grid=element_blank(),legend.position="none")
ggplotly()

# check individual 235399

ggplot(trackseg_argos_df[trackseg_argos_df$id==235399,], aes(lon, lat)) +
  geom_point(size=0.5, aes(col = as.factor(track_id ))) +
  geom_polygon(data=world_map,aes(x=long,y=lat,group=group))+
  coord_fixed(xlim=c(160,180), ylim=c(-56,-45))+
  theme_bw()+
  theme(panel.grid=element_blank())

ggplotly()

# check individual 46950
ggplot(trackseg_argos_df[trackseg_argos_df$id==46950,], aes(lon, lat)) +
  geom_point(size=0.5, aes(col = as.factor(track_id ))) +
  geom_polygon(data=world_map,aes(x=long,y=lat,group=group))+
  coord_fixed(xlim=c(160,180), ylim=c(-56,-45))+
  theme_bw()+
  theme(panel.grid=element_blank())

# check individual 197853
ggplot(trackseg_argos_df[trackseg_argos_df$id==197853,], aes(lon, lat)) +
  geom_point(size=0.5, aes(col = as.factor(track_id ))) +
  geom_polygon(data=world_map,aes(x=long,y=lat,group=group))+
  coord_fixed(xlim=c(160,180), ylim=c(-56,-45))+
  theme_bw()+
  theme(panel.grid=element_blank())

ggplotly()

# check individual 203575
ggplot(trackseg_argos_df[trackseg_argos_df$id==203575,], aes(lon, lat)) +
  geom_point(size=0.5, aes(col = as.factor(track_id ))) +
  geom_polygon(data=world_map,aes(x=long,y=lat,group=group))+
  coord_fixed(xlim=c(160,180), ylim=c(-56,-45))+
  theme_bw()+
  theme(panel.grid=element_blank())

```


```{r remove observation < 10 and locations in June}

min_obs <- 10 ## set the number of minimum obs acceptable
trackseg_argos_df <- trackseg_argos_df %>% group_by(track_id)
trackseg_argos_df_filt <- filter(trackseg_argos_df, n() >= min_obs)
table(trackseg_argos_df_filt$track_id)

# 50445 locations left

# remove locations in June 2023 
trackseg_argos_df_filt <- filter(trackseg_argos_df_filt,date < ymd_hms("2023-06-01 00:53:00",tz="GMT"))

str(trackseg_argos_df_filt)

#50276 locations left 

```

```{r}
ssm_df <- trackseg_argos_df_filt[,c(2:5,9)]

ssm_tdiff_hours_df <- ddply(ssm_df, ~track-id, function(d){
  d$time_diff_hours <- NA
  for (i in 2:nrow(d)){
    d$time_diff_hours[i] = as.numeric(difftime(d$date[i], d$date[i-1], units = "hours"))}
  return(d)
})

# mean time difference between locations (in hours)
mts <- aggregate(time_diff_hours~ track-id, ssm_tdiff_hours_df, mean)
mts #this is the mean time step
mean(mts$time_diff_hours)

# average 1 location every 3.6 hrs 

```

```{r}

ssm_df <- ssm_df[,c(5,1:4)]
colnames(ssm_df)[1] <- "id"
ssm_df <- data.frame(ssm_df)

# 50276 locations before running the ssm model 

```


2022 data 

n=8

Still transmitting
197853
235399
235404

Stop transmitting

208742
235400
235401
235402
235403

Removed 235399-8 as they gave WRONG move persistence estimates 

Pre ssm = 16021
Time step = 6h
after ssm = 5635

```{r ssm 2022}

# whales were tagged in 2022 only 


ssm_22 <- ssm_df%>%filter(date>"2022-07-01")

table(ssm_22$id)

#remove 215262 and 215258

ssm_22<- ssm_22  %>% subset (id != "215262-18"&
                            id != "215262-20"&
                            id!="215262-22"&
                            id!="215262-26"&
                            id!="215262-27"&  
                            id!="215262-30"&
                            id!="215262-33"&
                            id != "215258-9"&
                            id != "215258-11"&
                            id != "215258-13"&
                            id != "215258-14"&
                            id != "215258-16"&
                            id != "215258-18"&
                            id != "235399-8")

# check again 
length(unique(ssm_22$id))
table(ssm_df$id)
table(ssm_22$id)
# calculate time difference 


ssm_tdiff_hours_22_df <- ddply(ssm_22, ~ id, function(d){
  d$time_diff_hours <- NA
  for (i in 2:nrow(d)){
    d$time_diff_hours[i] = as.numeric(difftime(d$date[i], d$date[i-1], units = "hours"))}
  return(d)
})

# mean time difference between locations (in hours)
mts <- aggregate(time_diff_hours ~ id, ssm_tdiff_hours_22_df, mean)
mts #this is the mean time step
mean(mts$time_diff_hours) #3.03 hrs



# average time difference is 3 hrs 

fit_ssm_6_22 <- fit_ssm(ssm_22,vmax = 25,model="crw",time.step = 6,control = ssm_control(verbose=0))


print(fit_ssm_6_22)

summary(fit_ssm_6_22)
ssm_6_22_df <- grab(x=fit_ssm_6_22,what = "p")


plot(fit_ssm_6_22,what="p",ask=F)

# mpm 

mpm_6_22 <- fit_mpm(fit_ssm_6_22,model = "mpm",control = mpm_control(verbose=0))


mpm_6_22.df <- grab(mpm_6_22,what="f")

# join ssm and mpm 
ssm_mpm_22 <- aniMotum::join(fit_ssm_6_22,mpm_6_22,what="p",normalise=F)%>%
              as.data.frame()

ssm_mpm_22_a <- aniMotum::join(fit_ssm_6_22,mpm_6_22,what="p",normalise=F,group=F)%>%
                as.data.frame()

mean(ssm_mpm_22$g)


  ggplot(data.frame(ssm_mpm_22), aes(lon, lat)) +
    geom_point(size=1, aes(col = g)) +
    geom_polygon(data = world_map, aes(x=long, y=lat, group=group), fill="black") +
   coord_equal() + 
    coord_fixed(xlim=c(160,180), ylim=c(-56,-45))+
    theme_bw()+
    theme(panel.grid=element_blank())+
    sc
  ggplotly()
  
  
  ggplot(data.frame(ssm_mpm_22), aes(lon, lat))+
    geom_point(size=1,aes(color=as.factor(id))) +
    geom_polygon(data = world_map, aes(x=long, y=lat, group=group), fill="black") +
   coord_equal() + 
    coord_fixed(xlim=c(160,180), ylim=c(-56,-45))+
    theme_bw()+
    theme(panel.grid=element_blank())
  ggplotly()
  


 lapply(ssm_mpm_22,summary)
```

2021 data 

n=11

46633
46635
46950
46955
212499
212500
215258
215259
215261
215262
215263

Removed 46635-2 and 215258-9 and 215258-14 as they gave WRONG move persistence estimates due to insufficient data. Also, they were not in the study region 

Pre ssm = 16738
Time step = 6h
After ssm = 4959
Overall ssm = 5401 

```{r ssm 2021}

table(ssm_df$id)

ssm_21 <-ssm_df %>% subset(id == "46633-0"| 
                          id == "46635-0"| 
                            
                                 #id == "46635-2"| 
                            
                                 id == "46635-3"| 
                          
                                 id == "46950-0"|
                                  
                                 id == "46955-0"|
                                 
                                 id == "212499-0"|
                                   
                                 id == "212500-0"|
                                   
                                 id == "215258-0"|
                                 id == "215258-1"|
                                 id == "215258-3"|
                                 id == "215258-4"|
                                 #id == "215258-9"|
                                 id == "215258-11"|
                                 id == "215258-13"|
                                #id == "215258-14"|
                                 id == "215258-16"|
                                   
                                 id == "215259-0"|
                                 id == "215259-1"|
                                 id == "215259-2"|
                                 id == "215259-3"|
                                 id == "215259-5"|
                                   
                                 id == "215261-0"|
                                 id == "215261-2"|
                            
                                 id == "215263-0")

table(ssm_21$id)

# calculate time difference 
ssm_tdiff_hours_21_df <- ddply(ssm_21, ~ id, function(d){
  d$time_diff_hours <- NA
  for (i in 2:nrow(d)){
    d$time_diff_hours[i] = as.numeric(difftime(d$date[i], d$date[i-1], units = "hours"))}
  return(d)
})

# mean time difference between locations (in hours)
mts <- aggregate(time_diff_hours ~ id, ssm_tdiff_hours_21_df, mean)
mts #this is the mean time step
mean(mts$time_diff_hours)

# average time difference is 2.73 hrs


fit_ssm_6_21 <-  fit_ssm(ssm_21,vmax = 25,model="crw",time.step =6,control = ssm_control(verbose=0))
ssm_6_21.df <- fit_ssm_6_21%>%
               grab(.,what="p")%>%
               as.data.frame()



# change longitude to 0-360 
mpm_21<- ssm_6_21.df %>%
              select(id,date,lon,lat)%>%
               mutate(lon = lon %% 360) %>%
               fit_mpm(., model = "mpm")
mpm_21.df <- grab(x=mpm_21,what = "f") 

ssm_mpm_21 <- aniMotum::join(fit_ssm_6_21,mpm_21,what="p")%>%
              as.data.frame()

ggplot(data.frame(ssm_mpm_21), aes(lon, lat)) +
  geom_point(size=1, aes(col = g)) +
  geom_polygon(data = world_map, aes(x=long, y=lat, group=group), fill="black") +
 coord_equal() + 
  coord_fixed(xlim=c(160,180), ylim=c(-56,-45))+
  theme_bw()+
  theme(panel.grid=element_blank())+
  sc
ggplotly()


ggplot(data.frame(ssm_mpm_21), aes(lon, lat)) +
  geom_point(size=1,aes(color=as.factor(id))) +
  geom_polygon(data = world_map, aes(x=long, y=lat, group=group), fill="black") +
 coord_equal() + 
  coord_fixed(xlim=c(160,200), ylim=c(-56,-45))+
  theme_bw()+
  theme(panel.grid=element_blank())
 
ggplotly()


 lapply(ssm_mpm_21,summary)


```


Removed 215262-22 and 215262-27 as they gave WRONG move persistence estimates due to insufficient data. Also, they were not in the study region 

Pre ssm = 1248
Time difference = 5.99h
Time step = 11h
After ssm = 431


215262-18 

pre ssm = 22 
time difference = 9.07h
time step = 20h
after ssm = 11

```{r 2021 215262}


table(ssm_df$id)

ssm_215262 <- ssm_df %>% subset( id == "215262-0"|
                                 id == "215262-1"|
                                 id == "215262-2"|
                                 id == "215262-4"|
                                 id == "215262-6"|
                                 id == "215262-7"|
                                 id == "215262-9"|
                                 id == "215262-12"|
                                 id == "215262-13"|
                                 id == "215262-14"|
                                 id == "215262-15"|
                                 id == "215262-16"|
                               # id == "215262-18"|
                                id == "215262-20"|
                                #id =="215262-22"|
                                id == "215262-26"|
                                #id == "215262-27"|   
                                id == "215262-30"|
                                id == "215262-33")



# calculate time difference 
ssm_tdiff_hours_215262_df <- ddply(ssm_215262, ~ id, function(d){
  d$time_diff_hours <- NA
  for (i in 2:nrow(d)){
    d$time_diff_hours[i] = as.numeric(difftime(d$date[i], d$date[i-1], units = "hours"))}
  return(d)
})

# mean time difference between locations (in hours)
mts <- aggregate(time_diff_hours ~ id, ssm_tdiff_hours_215262_df, mean)
mts #this is the mean time step
mean(mts$time_diff_hours)


fit_ssm_215262 <- fit_ssm(ssm_215262,vmax = 25,model="crw",time.step =11,control = ssm_control(verbose=0))
ssm_215262_df <- grab(x=fit_ssm_215262,what = "p") 

#mpm
mpm_215262 <- fit_mpm(fit_ssm_215262,model="mpm",control = mpm_control(verbose = 0))
mpm_215262_df <- grab(x=mpm_215262,what = "f") 

ssm_mpm_215262 <- aniMotum::join(fit_ssm_215262,mpm_215262,what="p")%>%
              as.data.frame()


# fit 215262-18 separately 

ssm_215262_18 <- ssm_df%>% subset(id == "215262-18")

# calculate time difference 
ssm_tdiff_hours_215262_18_df <- ddply(ssm_215262_18, ~ id, function(d){
  d$time_diff_hours <- NA
  for (i in 2:nrow(d)){
    d$time_diff_hours[i] = as.numeric(difftime(d$date[i], d$date[i-1], units = "hours"))}
  return(d)
})

# mean time difference between locations (in hours)
mts <- aggregate(time_diff_hours ~ id, ssm_tdiff_hours_215262_18_df, mean)
mts #this is the mean time step
mean(mts$time_diff_hours)


# fit 215262-18
fit_ssm_215262_18 <- fit_ssm(ssm_215262_18,vmax = 25,model="crw",time.step =20,control = ssm_control(verbose=0))
ssm_215262_18_df <- grab(x=fit_ssm_215262_18,what = "p") 
#mpm
mpm_215262_18 <- fit_mpm(fit_ssm_215262_18,model="mpm",control = mpm_control(verbose = 0))
mpm_215262_df <- grab(x=mpm_215262_18,what = "f") 


ssm_mpm_215262_18 <- aniMotum::join(fit_ssm_215262_18,mpm_215262_18,what="p")%>%
              as.data.frame()


ssm_mpm_21 <- rbind(ssm_mpm_21,ssm_mpm_215262)
ssm_mpm_21 <- rbind(ssm_mpm_21,ssm_mpm_215262_18)


```

2020 data 

n=6

203571
203572
203573
203574
203575
205015

Pre ssm = 16111 (w/o 203573-1)
time difference = 1.81h
Time step = 6h
after ssm = 3750

cant get fit_mpm work with 203573-1 

pre ssm = 11 (203573-1)
time difference = 2.89h 
time step = 
after ssm =

```{r 2020}

table(ssm_df$id)

ssm_20 <- ssm_df %>% subset (id == "203571-0"|
                             id == "203571-1"|
                               
                            id == "203572-0"|
                            id == "203572-1"|
                              
                              id == "203573-0"|
                                  #id == "203573-1"|
                                    
                                  id == "203574-0"|
                                    
                                  id == "203575-0"|
                                  id == "203575-2"|
                              
                                  id == "205015-0")

# calculate time difference 
ssm_tdiff_hours_20_df <- ddply(ssm_20, ~ id, function(d){
  d$time_diff_hours <- NA
  for (i in 2:nrow(d)){
    d$time_diff_hours[i] = as.numeric(difftime(d$date[i], d$date[i-1], units = "hours"))}
  return(d)
})

# mean time difference between locations (in hours)
mts <- aggregate(time_diff_hours ~ id, ssm_tdiff_hours_20_df, mean)
mts #this is the mean time step
mean(mts$time_diff_hours)

fit_ssm_20<- fit_ssm(ssm_20, vmax=25, model="crw", time.step=6, control = ssm_control(verbose=0))
ssm_20.df <- grab(fit_ssm_20,what="p") 

mpm_20<- fit_mpm(fit_ssm_20,model="mpm",control = mpm_control(verbose = 0))
mpm_20_df <- grab(x=mpm_20,what = "f") 

ssm_mpm_20 <- aniMotum::join(fit_ssm_20,mpm_20,what="p")%>%
              as.data.frame()

ggplot(data.frame(ssm_mpm_20), aes(lon, lat)) +
  geom_point(size=1,aes(color=g)) +
  geom_polygon(data = world_map, aes(x=long, y=lat, group=group), fill="black") +
 coord_equal() + 
  coord_fixed(xlim=c(160,180), ylim=c(-56,-45))+
  theme_bw()+sc+
  theme(panel.grid=element_blank())
 
ggplotly()


 lapply(ssm_mpm_20,summary)


```



